---
date: "2025-10-13"
output:
  html_document:
    toc: false
    df_print: paged
---

```{r setup, include=FALSE}
meta <- rmarkdown::metadata
title_text <- "Cortinarius violaceus" # manually set since we removed YAML title
date_text <- meta$date
```

::: {style="text-align; margin-bottom: 1em;"}
# *`r title_text`*

**Date:** `r format(as.Date(date_text), "%B %d, %Y")`
:::

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c(
  "rinat", "dplyr", "sf", "ggplot2",
  "rnaturalearth", "rnaturalearthdata",
  "lubridate", "devtools"
)

# Install any that aren't already installed
installed <- packages %in% installed.packages()
if (any(!installed)) {
  install.packages(packages[!installed])
}

# Load all packages
lapply(packages, library, character.only = TRUE)
```

------------------------------------------------------------------------

## 1. Taxonomy & Nomenclature

-   **Scientific Name (current):** *Cortinarius violaceus*

-   **Synonyms / Former Names:** *Agaricus violaceus L.*, *Cortinarius violaceus (L.)* Fr., *Inoloma violaceum* (L. ex Fr.) Wunsche

-   **Common Name(s):** Violet cort, Violet webcap, Cap Gweog Fioled (Welsh)

-   **Taxonomic Authority / Citation:** (L.) Gray

-   **Higher Classification (Kingdom / Phylum / Class / Order / Family):** Fungi / Basidiomycota / Agaricomycetes / Agaricales / Cortinariaceae

------------------------------------------------------------------------

## 2. Morphology & Identification

-   **Fruiting Body Description:** The fruiting body takes the shape of a gilled mushroom that can be 4-10 cm wide and 6-15 cm tall. It has a violet cap and stipe and dark violet, blackish gills. It is generally textured with small hairs and scales.

-   **Cap / Pileus Characteristics:** The cap is typically 4-10 cm across and is somewhat velvety, roughened by many hairs and small scales. It tends to crack when dry and is colored a deep blackish violet to deep purple. Unlike many other Cortinarius species, the cap is not slimy.

-   **Gills / Pores / Spines / Teeth / Etc.:** The gills have an adnate attachment to the stem. They are covered by a cobwebby purple partial veil when young, but this disappears quickly. They begin as dark purple, become grayish to blackish, and end as rusty brown.

-   **Stipe / Stem / Attachment / Ring / Volva / Base:** The stipe is 6-15 cm long, with a base ranging from bulbous to slightly thickened, and has a partial veil. It is somewhat hairy and is not slimy. Over time, the stem darkens deeply, indicating age.

-   **Spore Print Color:** Rusty brown or cinnamon.

-   **Microscopic Features:** Spores are 11.5-14.5 x 7-9 µm and are ellipsoid to amygadaliform (almond-shaped), with a moderately verrucose (warty) surface texture. The cheilocystidia (edge of gills) and pleurocystidia (face of gills) are fusoid-ventricose (spindle-shaped, wider in the middle), with a long neck. It is the only Cortinarius species with clearly defined cystidia on its gills.

-   **Other Diagnostic Characters:** It is described by various sources as having a cedarwood smell, a sweet fragrance, or no distinct smell. It turns red when it comes into contact with potassium hydroxide (KOH).

-   **Similar Species / Potential Confusions:**

    -   Viscid violet cort (*Cortinarius iodes*)

    -   Gassy webcap (*Cortinarius traganus*)

    -   Silvery-violet cort (*Cortinarius alboviolaceus*)

    -   Amethyst deceiver (*Laccaria amethystina*)

    -   Blewit (*Clitocybe nuda*)

------------------------------------------------------------------------

## 3. Distribution & Habitat in Washington

-   **Range (counties or regions):** Temperate areas in the Northern Hemisphere (found in both North America and Europe).

-   **Geographic Distribution in WA:** Largely found to the west of the Cascade Mountains (in old-growth forests both inland and on the coast). Also sometimes found in the northeastern corner of the state (Pend Oreille County).

-   **Elevation Range:** There is no definitive elevation range for the fungus itself. However, the old-growth forests it favors occur from sea level to around 4,000 feet.

-   **Habitat Types / Substrate Preferences:** In Washington state, it is most often found growing in old-growth conifer forests, but it also grows in mixed woods. It grows on the ground and is often found near rotting logs.

-   **Host Associations:** It is more often reported with conifers in North America and deciduous forests of broadleaf trees in Europe.

```{r Distribution-map-WA, echo=FALSE, message=FALSE, warning=FALSE}

# 1. Get records by taxon_id (Hygrocybe singerii = 118316)
violetcort_data <- get_inat_obs(taxon_id = 124344, maxresults = 10000)

# 2. Convert to sf object
violetcort_sf <- violetcort_data %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# 3. Get WA state polygon
wa_state <- ne_states(country = "United States of America", returnclass = "sf") %>%
  filter(name == "Washington") %>%
  st_transform(st_crs(violetcort_sf))

# 4. Keep points inside WA
violetcort_wa <- violetcort_sf[st_within(violetcort_sf, wa_state, sparse = FALSE), ]

# 5. Plot
ggplot() +
  geom_sf(data = wa_state, fill = "gray95", color = "black") +
  geom_sf(data = violetcort_wa, color = "darkorchid4", alpha = 0.6, size = 2) +
  labs(
    title = "iNaturalist Observations of Cortinarius violaceus in Washington State",
    caption = "Data source: iNaturalist"
  ) +
  theme_classic()
```

------------------------------------------------------------------------

## 4. Seasonality & Phenology

-   **Months of Fruiting in WA:** June through December.

-   **Seasonal Notes:** Peaks in October and prefers moist soil.

```{r Seasonality-plot-WA, echo=FALSE, message=FALSE, warning=FALSE}

# Make sure your date column is parsed
violetcort_data <- violetcort_wa %>%
  mutate(
    observed_on = as.Date(observed_on),
    month_num = month(observed_on),
    month_label = month(observed_on, label = TRUE, abbr = TRUE)
  ) %>%
  filter(!is.na(month_num))

# Aggregate by month
monthly_counts <- violetcort_data %>%
  group_by(month_num, month_label) %>%
  summarize(observations = n(), .groups = "drop")

# Make sure months are in calendar order
monthly_counts <- monthly_counts %>%
  arrange(month_num)

#make proportional
monthly_counts <- monthly_counts %>%
  mutate(freq = observations / sum(observations))

# Plot
ggplot(monthly_counts, aes(x = month_num, y = freq)) +
  geom_area(fill = "orchid4", alpha = 0.5) +
  geom_line(color = "purple4", size = 1.2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = expression("Seasonal Pattern of " * italic("Cortinarius violaceus") * " (Relative Frequency)"),
    x = "Month",
    y = "Proportion of Observations"
  ) +
  theme_classic()
```

------------------------------------------------------------------------

## 5. Ecology & Role

-   **Ecological Function:** Forms ectomycorrhizal relationships with hardwoods or conifers.

-   **Substrate / Decomposition Role:** It is not known as a saprotrophic mushroom.

-   **Interactions:** As an ectomycorrhizal fungus, it enhances nutrient and water uptake for host trees while receiving carbohydrates in return.

-   **Environmental Sensitivity / Indicators:** Indicates healthy old-growth forests.

------------------------------------------------------------------------

## 6. Conservation, Abundance & Threats

-   **Abundance / Rarity in WA:** Commonly encountered in WA, but in small numbers.

-   **Conservation Status (state, federal, red list):** Not of concern.

-   **Threats / Pressures:** Currently, there are no specific threats to this species. However, it is assumed that when old-growth forests are impacted, this species will be as well.

-   **Management or Monitoring Notes:** Requires conservation of mature forest stands, avoidance of clearcutting and soil compaction, and maintenance of leaf litter, coarse woody debris (CWD), and natural hydrology.

------------------------------------------------------------------------

## 7. Use / Ethnobiology / Toxicology

-   **Edibility / Toxicity:** They are considered edible, but not choice, and so are not recommended for consumption. They may taste bitter when cooked.

-   **Traditional / Cultural Uses:** There are currently no documented uses for this species for culinary or medicinal uses.

-   **Medicinal / Bioactive Compounds:** It likely contains some compounds with anti-inflammatory and antioxidant properties, but more research is needed.

-   **Cautions / Warnings:** There are no serious warnings for consumption or use of this species. However, it has some lookalikes within the Cortinarius family that are toxic.

------------------------------------------------------------------------

## 8. References / Further Reading

-   Key Literature Sources:
    -   All That the Rain Promises, and More… (David Arora)
    -   Mushrooms Demystified (David Arora)
    -   Mushrooms of Cascadia (Noah Siegel and Christian Schwarz)\
-   Field Guides / Online Resources:
    -   <https://www.mushroomexpert.com/cortinarius_violaceus.html>
    -   <https://species.nbnatlas.org/species/NHMSYS0001479566>
    -   <https://burkeherbarium.org/imagecollection/taxon.php?Taxon=Cortinarius%20violaceus>
    -   <https://www.inaturalist.org/taxa/124344-Cortinarius-violaceus>
    -   <https://pmc.ncbi.nlm.nih.gov/articles/PMC10647524/>

------------------------------------------------------------------------

## 9. Images / Illustrations

-   **Fruiting Body Photos:**

    ![]()

    <p style="font-size:0.8em; text-align:center;">

    Figure 1: Five Cortinarius violaceus fruiting bodies in various stages of life.

    Photo credit: Federico Calledda

    </p>

    ![Underside / Hymenium]()

    <p style="font-size:0.8em; text-align:center;">

    Figure 2: A young specimen with some of its cobwebby partial veil still visible.

    Photo credit: Kavya Rao

    </p>

    ![Underside / Hymenium]()

    <p style="font-size:0.8em; text-align:center;">

    Figure 3: One of the authors (Kavya) and her partner posing with a lovely pair of violet webcaps!

    Photo credit: Kavya Rao

    </p>

-   **Habitat Photo:**

    ![Habitat]()

    <p style="font-size:0.8em; text-align:center;">

    Figure 4: Cortinarius violaceus growing among moss.

    Photo credit: Steve Trudell

    </p>

    ![Habitat]()

    <p style="font-size:0.8em; text-align:center;">

    Figure 5: Cortinarius violaceus growing on the forest floor.

    Photo credit: Steve Trudell

    </p>

-   **Microscopy Photo:**

    ![Microscopy]()

    <p style="font-size:0.8em; text-align:center;">

    Figure 6: Spores of Cortinarius Violaceus next to a scale.

    Photo credit: Michael Kuo

    </p>

------------------------------------------------------------------------

## 10. Other Notes & Observations

-   Field Notes / Observations:

    -   I (Kavya) have always found this fungus growing either by itself or in small groups of 3-5.

-   Questions / Uncertainties:

    -   The description of the smell is somewhat different between sources (cedar wood, vague sweetness, or no distinctive smell).

-   Fun facts

    -   Its fruiting body contains around 100 times more iron than most other fungal species.

------------------------------------------------------------------------
