---
date: "2025-10-13"
output:
  html_document:
    toc: false
    df_print: paged
---

```{r setup, include=FALSE}
meta <- rmarkdown::metadata
title_text <- "Rhizopogon vinicolor" # manually set since we removed YAML title
date_text <- meta$date
```

::: {style="text-align; margin-bottom: 1em;"}
# *`r title_text`*

**Date:** `r format(as.Date(date_text), "%B %d, %Y")`
:::

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c(
  "rinat", "dplyr", "sf", "ggplot2",
  "rnaturalearth", "rnaturalearthdata",
  "lubridate", "devtools"
)

# Install any that aren't already installed
installed <- packages %in% installed.packages()
if (any(!installed)) {
  install.packages(packages[!installed])
}

# Load all packages
lapply(packages, library, character.only = TRUE)
```

------------------------------------------------------------------------

## 1. Taxonomy & Nomenclature

-   **Scientific Name (current):** *Rhizopogon vinicolor*

-   **Synonyms / Former Names:** NA

-   **Common Name(s):** Wine-colored Pogie

-   **Taxonomic Authority / Citation:** Mycobank

-   **Higher Classification (Kingdom / Phylum / Class / Order / Family):\
    **Fungi -\> Dikarya -\> Basidiomycota -\> Agaricomycotina -\> Agaricomycetes -\> Agaricomycetidae -\> Boletales -\> Rhizopogonaceae -\> Rhizopogon -\> Rhizopogon Vinicolor

------------------------------------------------------------------------

## 2. Morphology & Identification

-   **Fruiting Body Description:**

    -   Size: 10 to 33 mm in diameter
    -   Shape: rounded or irregularly shaped
    -   Color: dusty white color and later red
    -   Texture: few rhizomorphs at the base

-   **Cap / Pileus Characteristics:\
    **Single layer, becomes wine red when exposed.

-   **Gills / Pores / Spines / Teeth / Etc.:**\
    Filled with basidiospores.

-   **Stipe / Stem / Attachment / Ring / Volva / Base:**\
    Base with few rhizomorphs.

-   **Spore Print Color:**\
    NA

-   **Microscopic Features (if available):** (spore size & shape, cystidia, etc.)

    -   Trama filled with basidiospores
    -   Basidiospores elliptical to ovid
    -   5-9 x 3-5 um in size
    -   Do not have a septum, smooth, and hyaline
    -   Mycelial cords are few in number of absent altogether

-   **Other Diagnostic Characters:**

    -   Odor: slightly fruity
    -   Bruising reactions: becomes vinaceous when exposed or bruised
    -   Color change: peridium becomes wine red when exposed. The body starts as a dirty color with tinges of yellow and staining pink, then becomes wine red.
    -   Texture: rubbery

-   **Similar Species / Potential Confusions:**

    -   *Rhizopogon vesiculosus,* and some species of *Amanita* (early in their life cycles)

------------------------------------------------------------------------

## 3. Distribution & Habitat in Washington

-   **Range (counties or regions):**\
    Clatlam, Snohomish, Kitsap, Pierce, Thurston, Lewis, Cowlitz, Clark, Skamania, Klickitat, Kittilas, Pend Oreille

-   **Geographic Distribution in WA:**\
    Western Coastal Washington and Central Cascades

-   **Elevation Range:**\
    Wide elevation range: sea level to 610m (Western Washington) and up to 1520 m (Central Cascades)

-   **Habitat Types / Substrate Preferences:** (forest type, soil, leaf litter, dead wood, etc.)

    Old growth Douglas fir stands, soil, roots

-   **Host Associations (if applicable):**

    Douglas fir

```{r Distribution-map-WA, echo=FALSE, message=FALSE, warning=FALSE}

# 1. Get records by taxon_id (Rhizopogon vinicolor =  384122)
rhizopogon_data <- get_inat_obs(taxon_id = 384122, maxresults = 10000)

# 2. Convert to sf object
rhizopogon_sf <- rhizopogon_data %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# 3. Get WA state polygon
wa_state <- ne_states(country = "United States of America", returnclass = "sf") %>%
  filter(name == "Washington") %>%
  st_transform(st_crs(rhizopogon_sf))

# 4. Keep points inside WA
rhizopogon_wa <- rhizopogon_sf[st_within(rhizopogon_sf, wa_state, sparse = FALSE), ]

# 5. Plot
ggplot() +
  geom_sf(data = wa_state, fill = "gray95", color = "black") +
  geom_sf(data = rhizopogon_wa, color = "purple", alpha = 0.6, size = 2) +
  labs(
    title = "iNaturalist Observations of Rhizopogon vinicolor in Washington State",
    caption = "Data source: iNaturalist"
  ) +
  theme_classic()
```

------------------------------------------------------------------------

## 4. Seasonality & Phenology

-   **Months of Fruiting in WA:**\
    The primary fruiting period is in the spring from February to July, but they can be found year round.

-   **Seasonal Notes:** (peak times, moisture dependence, etc.)\
    Fruiting season is triggered by rain. The best time to look is in the weeks after significant rain has saturated the forest floor. The fungus itself is drought-resistant and can help the host trees survive dry spells.

```{r Seasonality-plot-WA, echo=FALSE, message=FALSE, warning=FALSE}

# Make sure your date column is parsed
rhizopogon_data <- rhizopogon_wa %>%
  mutate(
    observed_on = as.Date(observed_on),
    month_num = month(observed_on),
    month_label = month(observed_on, label = TRUE, abbr = TRUE)
  ) %>%
  filter(!is.na(month_num))

# Aggregate by month
monthly_counts <- rhizopogon_data %>%
  group_by(month_num, month_label) %>%
  summarize(observations = n(), .groups = "drop")

# Make sure months are in calendar order
monthly_counts <- monthly_counts %>%
  arrange(month_num)

#make proportional
monthly_counts <- monthly_counts %>%
  mutate(freq = observations / sum(observations))

# Plot
ggplot(monthly_counts, aes(x = month_num, y = freq)) +
  geom_area(fill = "orchid4", alpha = 0.5) +
  geom_line(color = "purple4", size = 1.2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = expression("Seasonal Pattern of " * italic("Rhizopogon vinicolor") * " (Relative Frequency)"),
    x = "Month",
    y = "Proportion of Observations"
  ) +
  theme_classic()
```

------------------------------------------------------------------------

## 5. Ecology & Role

-   **Ecological Function:**\
    Ectomycorrhizal relationship with Douglas fir, saprotroph.

-   **Substrate / Decomposition Role:**\
    Saprotrophyic.

-   **Interactions:**

    They have a mutualistic relationship with Douglas firs. They breakdown leaf litter and rotting wood for nutrient availability for the Douglas Firs in exchange for carbohydrates.

    Western red-backed vole feeds on their fruiting bodies, the spores pass through their guts, and are deposited in fecal pellets. These spores remain viable and form associations with uninfected trees. 

-   **Environmental Sensitivity / Indicators:**

    If a forest is clear-cut and the dead wood is removed, *Rhizopogon vinicolor* stops fruiting, the vole population dies out, and newly planted trees fail to thrive. Tree growth is positively associated with the presence of *R. vinicolor*.

------------------------------------------------------------------------

## 6. Conservation, Abundance & Threats

-   **Abundance / Rarity in WA:**\
    Highly abundant in Douglas fir stands in Washington. 

-   **Conservation Status (state, federal, red list):**\
    Not of concern

-   **Threats / Pressures:**\
    Clear-cutting of old-growth Douglas fir stands

-   **Management or Monitoring Notes:**

    Preserve old-growth Douglas firs to maintain the relationship and presence between Douglas firs, voles, and *R. vinicolor*.

------------------------------------------------------------------------

## 7. Use / Ethnobiology / Toxicology

-   **Edibility / Toxicity:**\
    It is edible, but due to its small size, rubbery/foam-like texture, and mild flavor, some say it requires more effort in the kitchen than the results justify.

-   **Traditional / Cultural Uses:**\
    No known traditional or cultural uses.

-   **Medicinal / Bioactive Compounds:**\
    It has the ability to inhibit the growth of certain root pathogens like Phytophthora cinnamomi and Pythium debaryanum. 

-   **Cautions / Warnings:\
    **Advised not to eat unless you are 100% sure of the identity, as some species of *Amanita*, which are highly toxic and begin life in an egg-shaped sac, may look similar. 

------------------------------------------------------------------------

## 8. References / Further Reading

-   Key Literature Sources: 

    KRETZER, ANNETTE M., et al. “Patterns of vegetative growth and gene flow in rhizopogon vinicolor and R. vesiculosus (Boletales, Basidiomycota).” Molecular Ecology, vol. 14, no. 8, 11 May 2005, pp. 2259–2268, <https://doi.org/10.1111/j.1365-294x.2005.02547.x.> 

    Luoma DL, Frenkel RE, Trappe JM. Fruiting of hypogeous fungi in Oregon Douglas-fir forests: Seasonal and habitat variation. Mycologia. 1991;83(3):335-353. <doi:10.1080/00275514.1991.12026018>

    Zak, B. “Characterization and classification of mycorrhizae of Douglas fir. II. pseudotsuga menziesii + rhizopogon vinicolor.” Canadian Journal of Botany, vol. 49, no. 7, 1 July 1971, pp. 1079–1084, <https://doi.org/10.1139/b71-154.> 

-   Field Guides / Online Resources:

    Trappe, James, and Matt Trappe. “Specific Habitats: Seasonality and Habitat Associations of Some PNW Truffles.” Fruiting Patterns and Habitat Associations of PNW Truffles, 2007, www.natruffling.org/blurbs.htm#:\~:text=Name:%20Cystangium%20vesiculosum,middle%20elevations%20in%20the%20mountains. 

    “Wine-Colored Pogie (Rhizopogon Vinicolor).” iNaturalist, 13 Jan. 2024, www.inaturalist.org/taxa/384122-Rhizopogon-vinicolor#articles-tab. 

-   Herbarium / Specimen Records:

    WTU Herbarium, Burke Museum. “Rhizopogon Vinicolor.” Rhizopogon Vinicolor - Burke Herbarium Image Collection, University of Washington, burkeherbarium.org/imagecollection/taxon.php?Taxon=Rhizopogon+vinicolor. Accessed 23 Oct. 2025. 

------------------------------------------------------------------------

## 9. Images / Illustrations

-   **Fruiting Body Photos:**

    ![]()\

    <p style="font-size:0.8em; text-align:center;">

    Description; Photo credit: Photographer Name

    </p>

    ![]()\

    <p style="font-size:0.8em; text-align:center;">

    Description; Photo credit: Photographer Name

    </p>

-   **Habitat Photo:**

    ![]()\

    <p style="font-size:0.8em; text-align:center;">

    Description; Photo credit: Photographer Name

    </p>

-   **Microscopy Photo (if available):**

    ![]()\

    <p style="font-size:0.8em; text-align:center;">

    Description, Photo credit: Photographer Name

    </p>

------------------------------------------------------------------------

## 10. Other Notes & Observations

-   **Field Notes / Observations:** Some have observed its existence around pines and other firs besides the Doug fir.
-   **Questions / Uncertainties:**
    -   Uncertain about the elevation range 
    -   No cultural importance was written about, but it could exist, just not in the literature. 
-   **Fun facts**
    -   *R. vinicolor* can grow under significant water stress and remains active under drought conditions. This leads to increased establishment and survival rate of inoculated tree seedlings on harsh, dry sites. 

    -   *R. vinicolor* can enhance the net photosynthesis of Douglas-fir seedlings. 

    -   *R. vinicolor* is a valuable model organism for studying the population genetics of ectomycorrhizal fungi due to its clear association with Douglas-fir and its distinct genetic markers.

        \
        \
        \

------------------------------------------------------------------------
